---
title: "Philadelphia Housing Price Prediction"
subtitle: "A Predictive Modeling Framework to Support Property Tax Assessment"
author: "Daisy, Liz, Johnny, Parker"
date: "March 17, 2026"
format: 
  revealjs:
    theme: simple
    slide-number: true
    slide-level: 2
    chalkboard: true
    code-line-numbers: true
    incremental: false
    smaller: true
    scrollable: true
---

```{r setup}
#| echo: false
#| output: false
#| message: false
#| warning: false

library(sf)
library(tidyverse)
library(dplyr)
library(scales)
library(ggplot2)
library(caret)
library(knitr)
library(modelsummary)
library(broom)

phl_props_filtered <- readRDS("data/phl_props_filtered.rds")
phl_nhoods <- readRDS("data/phl_nhoods.rds")
model_df_cv <- readRDS("data/model_df_cv.rds")
cv_m3 <- readRDS("data/cv_m3.rds")
cv_m4 <- readRDS("data/cv_m4.rds")
cv_m5b <- readRDS("data/cv_m5b.rds")
cv_m7 <- readRDS("data/cv_m7.rds")
```



# Research Question & Motivation {visibility="hidden"}

## Research Question & Motivation

:::: {.columns}
::: {.column width="50%"}
### Research Question

- Background: Philadelphia is seeking to improve its **Automated Valuation Model**.
- Research question: How can residential sale prices prediction be **more accurate**?
- Analytical scope: The segment of residential **marketing** sales.
:::

::: {.column width="50%"}
### Motivation

- Property tax assessment depends on **credible estimates** of market value.
- Inaccurate valuation can lead to **unequal policy**
- Better prediction help **strengthen AVM**
:::
::::

---

# Data Overview {visibility="hidden"}


## Data Sources

:::: {.columns}
::: {.column width="33%"}
🏠

### Property Data

- Residential sales records
- Sale price
- Interior area
- Bedrooms / bathrooms
- Year built
- Property type
:::

::: {.column width="33%"}
🏘️

### Neighborhood Data

- ACS census variables
- Median income
- Poverty
- Education
- Demographic structure
:::

::: {.column width="33%"}
📍

### Spatial Data

- Transit access
- Park proximity
- Distance-based accessibility
- Other city or amenity measures
:::
::::

*Together, these sources capture property characteristics, neighborhood market conditions, and their location information.*

---

## Sample Definition and Data Cleaning

:::: {.columns}
::: {.column width="50%"}
### Sample Definition

- Property type: Single family & multi-family
- Sale years: 2023–2024
- Sale price: > $50,000
- Living area: > 0 sq ft
- Bedrooms: > 0
- State: Pennsylvania only
- Log-transformed sale price as outcome variable
:::

::: {.column width="50%"}
### Data Cleaning

- Removed records with missing or empty geometry
- Dropped missing values in key variables
- Filtered invalid coordinates
- Joined census tract data via spatial join
- Grouped sparse neighborhoods (n < 10) for cross-validation stability
:::
::::

---

*Final analytic sample: XX,XXX residential sales across Philadelphia, 2023–2024.*


# Key Visualizations {visibility="hidden"}

## Spatial Distribution of Residential Sale Prices

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
ggplot() +
  geom_sf(data = phl_nhoods, fill = "grey95", color = "white", linewidth = 0.3) +
  geom_sf(data = phl_props_filtered, aes(color = log_sale_price), size = 0.001) +
  scale_color_viridis_c(option = "plasma", name = "Log Sale Price") +
  theme_void() +
  labs(title = "Residential Sale Prices Across Philadelphia (2023–2024)")
```

---

## Property Features Associated with Sale Price

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
ggplot(phl_props_filtered, aes(x = total_livable_area, y = sale_price)) +
  geom_point(alpha = 0.3, size = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_continuous(labels = scales::comma) +
  theme_minimal() +
  labs(
    title = "Sale Price vs. Interior Area",
    x = "Total Livable Area (sq ft)",
    y = "Sale Price ($)"
  )
```

---

## Neighborhood Context and Market Value

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
ggplot(model_df_cv, aes(x = med_inc, y = log_sale_price)) +
  geom_point(alpha = 0.3, size = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  scale_x_continuous(labels = scales::dollar) +
  theme_minimal() +
  labs(
    title = "Log Sale Price vs. Median Household Income",
    x = "Median Household Income ($)",
    y = "Log Sale Price"
  )
```

---

# Model Comparison {visibility="hidden"}

## Layered Modeling Strategy

| Layer | Variables Added |
|-------|----------------|
| M3: Structure + Age | Living area, bedrooms, bathrooms, age |
| M4: + Census | Median income, poverty, education, transit share |
| M5b: + Spatial | Transit access, trees, school distance, dist. to CBD |
| M7: + Neighborhood FE | Neighborhood fixed effects |

*Each layer adds more information about where a home is and who lives nearby.*

---

## Comparative Model Performance

```{r}
#| eval: true
#| echo: false
data.frame(
  Model = c(
    "M3: Structure + Age",
    "M4: + Census",
    "M5b: + Spatial + Year/Type",
    "M7: + Neighborhood FE"
  ),
  RMSE = c(cv_m3$results$RMSE, cv_m4$results$RMSE,
           cv_m5b$results$RMSE, cv_m7$results$RMSE),
  MAE = c(cv_m3$results$MAE, cv_m4$results$MAE,
          cv_m5b$results$MAE, cv_m7$results$MAE),
  Rsquared = c(cv_m3$results$Rsquared, cv_m4$results$Rsquared,
               cv_m5b$results$Rsquared, cv_m7$results$Rsquared)
) %>%
  kable(digits = 4, caption = "10-Fold Cross-Validation Results")
```

---

# Main Findings {visibility="hidden"}

## Final Model Accuracy

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
ggplot(cv_m7$pred, aes(x = obs, y = pred)) +
  geom_point(alpha = 0.3, size = 0.5, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Predicted vs. Actual Log Sale Price (M7, 10-fold CV)",
    x = "Actual Log Sale Price",
    y = "Predicted Log Sale Price"
  )
```

---

## Most Influential Predictors

## Areas of Persistent Prediction Error

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
model_df_cv %>%
  mutate(
    residual_m7 = log_sale_price - predict(
      cv_m7, newdata = model_df_cv
    )
  )

model_df_cv %>%
  group_by(neighborhood_cv) %>%
  summarise(mean_abs_error = mean(abs(residual_m7), na.rm = TRUE)) %>%
  arrange(desc(mean_abs_error)) %>%
  head(10) %>%
  kable(digits = 4, caption = "Top 10 Neighborhoods with Highest Prediction Error")
```

---

# Policy Recommendations {visibility="hidden"}

## Risks and Limitations for OPA

:::: {.columns}
::: {.column width="50%"}
### Risks

- Neighborhood fixed effects may encode historical bias
- One citywide model may not fit all submarkets equally
- Rapidly gentrifying areas are harder to predict
:::

::: {.column width="50%"}
### Limitations

- Missing data excluded from model
- Outlier removal may drop legitimate sales
- Model trained on 2023–2024 only
:::
::::

---

## Recommendations for Strengthening the AVM

- 
- 
- 
- 


