---
title: "Philadelphia Housing Price Prediction"
subtitle: "A Predictive Modeling Framework to Support Property Tax Assessment"
author: "Daisy, Liz, Johnny, Parker"
date: "March 17, 2026"
project:
  type: default
format:
  revealjs:
    theme: simple
    slide-number: true
    slide-level: 2
    chalkboard: true
    code-line-numbers: true
    incremental: false
    smaller: true
    scrollable: true
---

```{r setup}
#| echo: false
#| output: false
#| message: false
#| warning: false

library(sf)
library(tidyverse)
library(dplyr)
library(scales)
library(ggplot2)
library(caret)
library(knitr)
library(modelsummary)
library(broom)
library(ggtext)

acs_data <- readRDS("data/acs_data.rds")
model_df_fe <- readRDS("data/model_df_cv.rds")

phl_props_filtered <- readRDS("data/phl_props_filtered.rds")
phl_nhoods         <- readRDS("data/phl_nhoods.rds")
model_df_cv        <- readRDS("data/model_df_cv.rds")
cv_m3              <- readRDS("data/cv_m3.rds")
cv_m4              <- readRDS("data/cv_m4.rds")
cv_m5b             <- readRDS("data/cv_m5b.rds")
cv_m7              <- readRDS("data/cv_m7.rds")
m3                 <- readRDS("data/m3.rds")
m4                 <- readRDS("data/m4.rds")
m5b                <- readRDS("data/m5b.rds")
m7                 <- readRDS("data/m7.rds")
```


## Research Question & Motivation

:::: {.columns}
::: {.column width="50%"}
### Research Question

- Background: Philadelphia is seeking to improve its *Automated Valuation Model*.
- Research question: How can residential sale prices prediction be *more accurate* spagial and demographic?
- Analytical scope: The segment of residential *marketing* sales.
:::

::: {.column width="50%"}
### Motivation

- Property tax assessment depends on *credible estimates* of market value.
- Inaccurate valuation can lead to *unequal policy*
- Better prediction help *strengthen AVM*
:::
::::

---

## Data Sources

:::: {.columns}
::: {.column width="33%"}

### Property Data

- Residential sales records
- Sale price
- Interior area
- Bedrooms / bathrooms
- Year built
- Property type
:::

::: {.column width="33%"}

### Neighborhood Data

- ACS census variables
- Median income
- Poverty
- Education
- Demographic structure
:::

::: {.column width="33%"}

### Spatial Data

- Transit access
- Park proximity
- Distance-based accessibility
- Other city or amenity measures
:::
::::

*Together, these sources capture property characteristics, neighborhood market conditions, and their location information.*

---

## Data Source & Cleaning

Source: OPA Properties Public Dataset - 2023-2024 Residential Sales, Philadelphia

<br>

:::: {.columns}

::: {.column width="33%"}
*Filter 1: Basic Info*

- Residential: Single Family & Multi Family
- Years: sales in 2023 & 2024
- State: PA only
:::

::: {.column width="33%"}
*Filter 2: Implausible Values*

- Living area: > 0 sq ft
- Bedrooms: > 0
- *Sale price: > $50,000*
:::

::: {.column width="33%"}
*Filter 3: Missing Values*

- Sale price
- Living area
- Bedrooms & bathrooms
- All model variables
:::

::::

*23372 residential sales across Philadelphia, 2023-2024.*

---

## Spatial Distribution of Residential Sale Prices
```{r}
#| echo: false
#| message: false
#| warning: false

phl_props_filtered <- readRDS("data/phl_props_filtered.rds")
phl_nhoods         <- readRDS("data/phl_nhoods.rds")

ggplot() +
  geom_sf(data = phl_nhoods, fill = "grey95", color = "white", linewidth = 0.3) +
  geom_sf(data = phl_props_filtered, aes(color = log_sale_price), size = 0.001, alpha = 0.6) +
  scale_color_viridis_c(option = "mako", name = "Log Sale\nPrice", direction = -1) +
  annotate("text",
           x = 2748000, y = 225000,
           label = "High-price zones:\nCenter City & Chestnut Hill",
           color = "#0d0887", size = 2.8, hjust = 1, vjust = 0, lineheight = 1.5,
           fontface = "bold") +
  annotate("text",
           x = 2748000, y = 214000,
           label = "Low-price zone:\nNorth & Northeast Philadelphia",
           color = "#a8e6cf", size = 2.8, hjust = 1, vjust = 0, lineheight = 1.5,
           fontface = "bold") +
  theme_void() +
  theme(
    plot.title      = element_text(size = 14, face = "bold"),
    legend.title    = element_text(size = 10),
    legend.text     = element_text(size = 9),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  labs(title = "Residential Sale Prices Exhibit Strong Spatial Clustering",
       subtitle = "Residential Sale Prices Across Philadelphia (2023-2024)")
```

---

## Properties Sold by Census Tract
```{r}
#| echo: false
#| message: false
#| warning: false

prop_summary <- phl_props_filtered %>%
  st_drop_geometry() %>%
  group_by(GEOID) %>%
  summarise(
    mean_price   = mean(sale_price, na.rm = TRUE),
    n_properties = n()
  )

tract_analysis <- acs_data %>%
  left_join(prop_summary, by = "GEOID")

ggplot(tract_analysis) +
  geom_sf(aes(fill = n_properties)) +
  scale_fill_viridis_c(option = "mako", name = "Number of\nProperties", direction = -1) +
  theme_void() +
  theme(
    plot.title      = element_text(size = 14, face = "bold"),
    legend.title    = element_text(size = 10),
    legend.text     = element_text(size = 9),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  labs(
    title    = "Properties Sold by Census Tract",
    subtitle = "2023-2024 Residential Sales, Philadelphia"
  )
```

Key takeaways:

- Sales are unevenly distributed across census tracts, with market activity concentrated in a limited number of neighborhoods and much thinner transaction coverage in others.

## Spatial Autocorrelation of Sale Prices
```{r}
#| echo: false
#| message: false
#| warning: false

library(spdep)
phl_props_filtered <- phl_props_filtered %>%
  filter(!st_is_empty(geometry))

# Build spatial weights matrix directly from sf object
lw <- nb2listw(knn2nb(knearneigh(phl_props_filtered, k = 5)), style = "W")

# Global Moran's I
moran_global <- moran.test(phl_props_filtered$log_sale_price, listw = lw, zero.policy = TRUE)

# Local Moran's I (LISA)
lisa <- localmoran(phl_props_filtered$log_sale_price, listw = lw, zero.policy = TRUE)

# Classify LISA clusters
m <- mean(phl_props_filtered$log_sale_price)
phl_props_filtered <- phl_props_filtered %>%
  mutate(
    local_I      = lisa[, "Ii"],
    local_pval   = lisa[, "Pr(z != E(Ii))"],
    lag_logprice = lag.listw(lw, log_sale_price),
    lisa_cluster = case_when(
      local_pval < 0.05 & log_sale_price >  m & lag_logprice >  mean(lag_logprice) ~ "High-High",
      local_pval < 0.05 & log_sale_price <  m & lag_logprice <  mean(lag_logprice) ~ "Low-Low",
      local_pval < 0.05 & log_sale_price >  m & lag_logprice <= mean(lag_logprice) ~ "High-Low",
      local_pval < 0.05 & log_sale_price <= m & lag_logprice >  mean(lag_logprice) ~ "Low-High",
      TRUE ~ "Not Significant"
    )
  )

# LISA map
lisa_map <- ggplot(phl_props_filtered) +
  geom_sf(aes(color = lisa_cluster), size = 0.4, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "High-High"       = "#d73027",
      "Low-Low"         = "#4575b4",
      "High-Low"        = "#fc8d59",
      "Low-High"        = "#91bfdb",
      "Not Significant" = "#d9d9d9"
    ),
    name = "LISA Cluster"
  ) +
  labs(
    title = "Similar-Priced Properties Cluster Together Across Philadelphia",
    subtitle = paste0("Moran's I = ", round(moran_global$estimate[1], 3), ", p < 0.001")
  ) +
  theme_void() +
  theme(
    plot.title      = element_text(size = 12, face = "bold"),
    plot.subtitle   = element_text(size = 10),
    plot.caption    = element_markdown(size = 8, lineheight = 1.5), 
    plot.background = element_rect(fill = "white", color = NA)
  )

lisa_map
```

Key takeaways:

<p style="font-size:0.8em; line-height:1.6;">
*<span style='color:#d73027'>High-High</span>* clusters concentrate in *<span style='color:#d73027'>Center City and Chestnut Hill</span>*, with smaller clusters in Far Northeast, around UPenn, and the far southern tip.<br>
*<span style='color:#4575b4'>Low-Low</span>* clusters dominate *<span style='color:#4575b4'>North and West Philadelphia</span>*.<br>
The <span style='color:#999999'>absence of significant autocorrelation at the boundary</span> suggests an abrupt spatial discontinuity.

---

## Model Performance Improves with Each Layer
```{r}
#| echo: false
#| message: false
#| warning: false

library(kableExtra)

data.frame(
  Model = c(
    "M3: Structural Only",
    "M4: + Census",
    "M5b: + Spatial",
    "M7: + Neighborhood FE"
  ),
  RMSE_log = c(
    cv_m3$results$RMSE,
    cv_m4$results$RMSE,
    cv_m5b$results$RMSE,
    cv_m7$results$RMSE
  ),
  Approx_Percent_Error = 100 * (exp(c(
    cv_m3$results$RMSE,
    cv_m4$results$RMSE,
    cv_m5b$results$RMSE,
    cv_m7$results$RMSE
  )) - 1),
  MAE_log = c(
    cv_m3$results$MAE,
    cv_m4$results$MAE,
    cv_m5b$results$MAE,
    cv_m7$results$MAE
  ),
  Rsquared = c(
    cv_m3$results$Rsquared,
    cv_m4$results$Rsquared,
    cv_m5b$results$Rsquared,
    cv_m7$results$Rsquared
  )
) %>%
  arrange(RMSE_log) %>%
  kable(
    digits    = c(0, 4, 1, 4, 4),
    caption   = "10-Fold Cross-Validation Results",
    col.names = c("Model", "RMSE (log)", "Approx. Error (%)", "MAE (log)", "R-squared")
  ) %>%
  kable_styling(
    font_size        = 16,
    full_width       = FALSE,
    bootstrap_options = c("striped", "hover")
  ) %>%
  row_spec(1, bold = TRUE, background = "#e8f4f8")  # highlight the best model
```

Bottom line: Model performance improves at each stage, with the largest gain from adding *neighborhood context*. The final model with neighborhood fixed effects performs best, though prediction error remains meaningful.


## Most Influential Predictors
```{r}
#| echo: false
#| message: false
#| warning: false

library(broom)

tidy(m7) %>%
  filter(
    !str_detect(term, "neighborhood|Intercept"),
    p.value < 0.05
  ) %>%
  mutate(
    term = case_when(
      term == "log_area"                               ~ "Living Area (log)",
      term == "number_of_bedrooms"                     ~ "Bedrooms",
      term == "number_of_bathrooms"                    ~ "Bathrooms",
      term == "age_c"                                  ~ "Age",
      term == "I(age_c^2)"                             ~ "Age-squared",
      term == "med_inc"                                ~ "Median Income",
      term == "poverty_rate"                           ~ "Poverty Rate",
      term == "transit_share"                          ~ "Transit Commute Share",
      term == "edu_ba_share"                           ~ "BA+ Education Share",
      term == "burden_renter30_share"                  ~ "Renter Cost Burden",
      term == "transit_500ft"                          ~ "Transit Stops (500ft)",
      term == "n_trees_500ft"                          ~ "Street Trees (500ft)",
      term == "school_knn_3"                           ~ "Distance to Schools",
      term == "dist_core_mi"                           ~ "Distance to Center City",
      term == "sale_year2024"                          ~ "Sale Year 2024",
      term == "category_code_descriptionSINGLE FAMILY" ~ "Single Family",
      TRUE ~ term
    ),
    direction  = if_else(estimate > 0, "Positive", "Negative"),
    term_label = if_else(
      term %in% c("Bathrooms", "BA+ Education Share",
                  "Poverty Rate", "Distance to Center City"),
      paste0("*", term, "*"),
      term
    )
  ) %>%
  arrange(desc(abs(estimate))) %>%
  ggplot(aes(x = reorder(term_label, abs(estimate)), y = estimate, fill = direction)) +
  geom_col() +
  geom_errorbar(aes(ymin = estimate - 1.96 * std.error,
                    ymax = estimate + 1.96 * std.error),
                width = 0.3, color = "grey40") +
  scale_fill_manual(values = c("Positive" = "#d73027", "Negative" = "#4575b4")) +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.title      = element_text(size = 13, face = "bold"),
    plot.background = element_rect(fill = "white", color = NA),
    legend.title    = element_blank(),
    axis.text.y     = element_markdown()  # enables bold via *
  ) +
  labs(
    title = "Most Influential Predictors in M7",
    x     = NULL,
    y     = "Coefficient Estimate (log scale)"
  )
```

Key Takeaways:

- Living area is the strongest single predictor

- Socioeconomics outweigh amenities, confirming that *who lives nearby* matters more than *what's nearby*.

- Bathrooms signal overall quality.


## Final Model Data Errors

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
ggplot(cv_m7$pred, aes(x = obs, y = pred)) +
  geom_ribbon(aes(ymin = obs - 0.3, ymax = obs + 0.3),
              fill = "gray", alpha = 0.15) +
  geom_point(alpha = 0.3, size = 0.5, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  annotate("text", x = 13.5, y = 15,
         label = "+/- 0.3 log units (~35%)", 
         color = "gray", size = 4, hjust = 0, vjust = 0, fontface = "bold") +
  theme_minimal() +
  labs(
    title = "Predicted vs. Actual Log Sale Price (M7, 10-fold CV)",
    x     = "Actual Log Sale Price",
    y     = "Predicted Log Sale Price"
  )
```

::: {.callout-note}
M7 predicts mid-range prices well, but systematically *overestimates low-price properties* and struggles with *luxury outliers*: consistent with limited training data at the extremes.
:::

## Final Model Spatial Errors

```{r}
#| echo: false
#| message: false
#| warning: false

# Load necessary data
model_df_fe <- model_df_fe %>%
  mutate(
    fitted_m7    = fitted(m7),
    residuals_m7 = residuals(m7),
    error        = exp(log_sale_price) - exp(fitted_m7)
  )

resid_sf <- phl_props_filtered %>%
  left_join(
    dplyr::select(model_df_fe, pin, residuals_m7, error),
    by = "pin"
  ) %>%
  filter(!is.na(residuals_m7))

# Convert log residuals to dollar scale for interpretability
resid_sf <- resid_sf %>%
  mutate(error = exp(log_sale_price) - exp(log_sale_price - residuals_m7))

ggplot() +
  geom_sf(data = phl_nhoods, fill = "grey95", color = "white", linewidth = 0.3) +
  geom_sf(data = resid_sf, aes(color = residuals_m7), size = 0.4, alpha = 0.7) +
  scale_color_gradient2(
    low      = "#4575b4",
    mid      = "white",
    high     = "#d73027",
    midpoint = 0,
    name     = "Residual\n(log scale)"
  ) +
  theme_void() +
  theme(
    plot.title      = element_text(size = 13, face = "bold"),
    plot.subtitle   = element_text(size = 10),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  labs(
    title    = "Spatial Distribution of Prediction Errors (M7)",
    subtitle = "Red = over-predicted, Blue = under-predicted"
  )
```

The residuals from m7 are relatively dispersed rather than forming large, continuous clusters, which suggests that the final model captures much of the broad spatial structure in Philadelphia housing market. 
At the same time, some localized market variation is still not fully explained.


## Summary

**Model Accuracy** RMSE = 0.422 (log scale), R-squared = 0.627 (10-fold CV)

**Top Predictors**

- Neighborhood fixed effects (largest single gain: R-squared 0.582 -> 0.627)
- Living area (coef = 0.538): strongest structural predictor
- BA+ education share (coef = 0.428) & poverty rate (coef = -0.146): socioeconomics outweigh physical amenities

**Key Recommendations**

- Current AVM likely undervalues transit-accessible and gentrifying properties
- Sub-models by neighborhood type would improve accuracy in rapidly changing markets
- Flag high-residual properties for manual review before formal tax assessment

## Thank You

**Data Sources**

- City of Philadelphia, Office of Property Assessment. OPA Properties Public Dataset. https://opendataphilly.org
- U.S. Census Bureau. American Community Survey 5-Year Estimates, 2023. Via tidycensus R package.
- SEPTA. Transit Stops, Spring 2025. OpenDataPhilly.
- City of Philadelphia, Parks & Recreation. Street Tree Inventory, 2025. OpenDataPhilly.
- School District of Philadelphia. Schools Dataset. OpenDataPhilly.
- City of Philadelphia. Neighborhood Boundaries. OpenDataPhilly.
