---
title: "Philadelphia Housing Model - Technical Appendix"
author: "Daisy Yuan"
format: 
  html:
    code-fold: show
    toc: true
    toc-location: left
    theme: cosmo
editor: 
  markdown: 
    wrap: sentence
---

**This is exactly where spatial features come in!**

------------------------------------------------------------------------

## Load necessary libraries

```{r}
library(tidyverse)
library(sf)
library(tidycensus)
library(tigris)
library(MASS)
library(dplyr)
library(scales)
library(ggplot2)
library(caret)
library(car)
library(knitr)
library(readr)
library(patchwork)
```

::: callout-note
## Today's Goal

By the end of class, you'll build models that: - Incorporate spatial relationships - Account for neighborhood effects\
- Achieve much better prediction accuracy - Help you understand what drives housing prices
:::

------------------------------------------------------------------------

## Phase1: Data Preparation

### Step 1: Load and Clean Philadelphia Sales Data

#### Load data

```{r}
opa_path <- "C:/Users/10133/Desktop/UPENN/MUSA5080_PPA/Portfolio/lab_3/opa_properties_public.geojson"
opa <- st_read(opa_path, quiet = FALSE)
```

#### Filter data

```{r}
# Filter Primary data by basic definition
opa_filtered <- opa %>%
  filter(
    category_code_description %in% c("SINGLE FAMILY", "MULTI FAMILY"),
    year(as.Date(sale_date)) %in% c(2023, 2024),
    mailing_city_state == "PHILADELPHIA PA",
    !is.na(sale_price),
    !is.na(total_livable_area) & total_livable_area > 0,
    !is.na(number_of_bedrooms) & number_of_bedrooms > 0
  )
```

#### Clean Price Outliers

```{r}
# Plot the data to see distribution of sale_price
ggplot(opa_filtered, aes(x = sale_price)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  scale_x_log10(labels = scales::dollar) + 
  labs(title = "Distribution of Sale Price (Log Scale)",
       x = "Sale Price (log)",
       y = "Count") +
  theme_minimal()
# We found that part of the distribution looks similar to normal distribution

# Check at which threshold the data are the most similar to normal distribution
thresholds <- c(1000, 5000, 10000, 20000, 40000, 50000)

map_dfr(thresholds, function(t) {
  d <- opa_filtered %>% filter(sale_price > t) %>% pull(sale_price)
  tibble(
    threshold = t,
    n = length(d),
    skewness = (mean(log(d)) - median(log(d))) / sd(log(d)),  # Pearson's skewness
    mean_log = mean(log(d)),
    sd_log = sd(log(d))
  )
})
# We can see the skewness of threshold 50000 stay the most closely to 0 at 0.0023
# It's obvious that 50000 makes more sense

# we want to make the least errors in calculating so unit_price matters too
#opa_filtered <- opa_filtered %>%
  #mutate(unit_price = sale_price / total_livable_area)

opa_filtered <- opa_filtered %>%
  filter(sale_price > 50000)

```

**Explanation of outliers identification methodology:**  - We observed a cluster of extremely low sale_price values that are unlikely to represent arm’s length market transactions.
- We examined the raw and log-transformed sale_price distributions and tested several lower-bound thresholds to assess sensitivity.
- We selected 50,000 as the cutoff because it removes the most implausible low-price records while preserving the overall distribution shape and a sufficient sample size.

#### Non-market sales

```{r}
# Market rate and sales value
ggplot(opa_filtered, aes(x = market_value, y = sale_price)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  scale_x_log10(labels = scales::dollar) +
  scale_y_log10(labels = scales::dollar) +
  labs(title = "Market Value vs Sale Price",
       x = "Market Value (log)",
       y = "Sale Price (log)",
       caption = "Red line = perfect 1:1 relationship") +
  theme_minimal()

# Check the ratio distribution
opa_filtered <- opa_filtered %>%
  mutate(price_ratio = sale_price / market_value)
summary(opa_filtered$price_ratio)

# Quantile-based cutoffs
opa_filtered <- opa_filtered %>%
  filter(price_ratio >= quantile(price_ratio, 0.01, na.rm = TRUE),
         price_ratio <= quantile(price_ratio, 0.99, na.rm = TRUE))
```

**Explanation of non-market rate sales identification methodology:**  - We compared sale_price to market_value and found a strong overall alignment between the two.
- We examined the price_ratio (sale_price divided by market_value) and identified extreme values, including a maximum around 76, which likely reflects non market transactions or data issues.
- Because these extremes can distort the distribution, we used quantiles to keep only the statistically dominant range of the data.
- Based on this inspection, the 0.01 and 0.99 quantiles fall within an acceptable error range, so we filtered observations to this interval.

#### Adjust variables

```{r}
# Add basic variables
opa_filtered <- opa_filtered %>%
  mutate(age = year(as.Date(sale_date)) - as.numeric(year_built)) %>%
  mutate(bb_ratio = number_of_bathrooms / number_of_bedrooms)

# Pick useful variables
opa_var <- opa_filtered %>%
  dplyr::select(
    objectid, geometry, 
    sale_date, sale_price, market_value, building_code_description_new,
    total_livable_area, number_of_bedrooms, number_of_bathrooms, bb_ratio,
    number_stories, garage_spaces, 
    central_air, quality_grade, view_type,
    interior_condition, exterior_condition, age, 
    off_street_open, zip_code, census_tract, zoning, owner_1,
    category_code_description, fireplaces
  )

```

### Step 2: Load and Clean Secondary Data


## Phase2: Feature Engineering

### Step 1: what

```{r}
opa_path <- "C:/Users/10133/Desktop/UPENN/MUSA5080_PPA/Portfolio/lab_3/opa_properties_public.geojson"
opa <- st_read(opa_path, quiet = FALSE)
```

------------------------------------------------------------------------
