---
title: "Philadelphia Housing Model - Technical Appendix"
author: "Daisy Yuan"
format: 
  html:
    code-fold: show
    toc: true
    toc-location: left
    theme: cosmo
editor: 
  markdown: 
    wrap: sentence
---

**This is exactly where spatial features come in!**

------------------------------------------------------------------------

## Load necessary libraries

```{r}
library(tidyverse)
library(sf)
library(tidycensus)
library(tigris)
library(MASS)
library(dplyr)
library(scales)
library(ggplot2)
library(caret)
library(car)
library(knitr)
library(readr)
library(patchwork)
```

::: callout-note
## Today's Goal

By the end of class, you'll build models that: - Incorporate spatial relationships - Account for neighborhood effects\
- Achieve much better prediction accuracy - Help you understand what drives housing prices
:::

------------------------------------------------------------------------

## Phase1: Data Preparation

### Step 1: Load and Clean Philadelphia Sales Data

#### Load data

```{r}
opa_path <- "C:/Users/10133/Desktop/UPENN/MUSA5080_PPA/Portfolio/lab_3/opa_properties_public.geojson"
opa <- st_read(opa_path, quiet = FALSE)
```

#### Filter data

```{r}
# Filter Primary data by basic definition
opa_filtered <- opa %>%
  filter(
    category_code_description %in% c("SINGLE FAMILY", "MULTI FAMILY"),
    year(as.Date(sale_date)) %in% c(2023, 2024),
    mailing_city_state == "PHILADELPHIA PA",
    !is.na(sale_price),
    !is.na(total_livable_area) & total_livable_area > 0,
    !is.na(number_of_bedrooms) & number_of_bedrooms > 0
  )
```

#### Clean Price Outliers

```{r}
# Plot the data to see distribution of sale_price
ggplot(opa_filtered, aes(x = sale_price)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  scale_x_log10(labels = scales::dollar) + 
  labs(title = "Distribution of Sale Price (Log Scale)",
       x = "Sale Price (log)",
       y = "Count") +
  theme_minimal()
# We found that part of the distribution looks similar to normal distribution

# Check at which threshold the data are the most similar to normal distribution
thresholds <- c(1000, 5000, 10000, 20000, 40000, 50000)

map_dfr(thresholds, function(t) {
  d <- opa_filtered %>% filter(sale_price > t) %>% pull(sale_price)
  tibble(
    threshold = t,
    n = length(d),
    skewness = (mean(log(d)) - median(log(d))) / sd(log(d)),  # Pearson's skewness
    mean_log = mean(log(d)),
    sd_log = sd(log(d))
  )
})
# We can see the skewness of threshold 50000 stay the most closely to 0 at 0.0023
# It's obvious that 50000 makes more sense

# we want to make the least errors in calculating so unit_price matters too
#opa_filtered <- opa_filtered %>%
  #mutate(unit_price = sale_price / total_livable_area)

opa_filtered <- opa_filtered %>%
  filter(sale_price > 50000)

```

**Explanation of outliers identification methodology:**  - We observed a cluster of extremely low sale_price values that are unlikely to represent arm’s length market transactions.
- We examined the raw and log-transformed sale_price distributions and tested several lower-bound thresholds to assess sensitivity.
- We selected 50,000 as the cutoff because it removes the most implausible low-price records while preserving the overall distribution shape and a sufficient sample size.

#### Non-market sales

```{r}
# Market rate and sales value
ggplot(opa_filtered, aes(x = market_value, y = sale_price)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  scale_x_log10(labels = scales::dollar) +
  scale_y_log10(labels = scales::dollar) +
  labs(title = "Market Value vs Sale Price",
       x = "Market Value (log)",
       y = "Sale Price (log)",
       caption = "Red line = perfect 1:1 relationship") +
  theme_minimal()

# Check the ratio distribution
opa_filtered <- opa_filtered %>%
  mutate(price_ratio = sale_price / market_value)
summary(opa_filtered$price_ratio)
# The ratios are crazy with the max at 76, which makes no sense

# Quantile-based cutoffs
opa_filtered <- opa_filtered %>%
  filter(price_ratio >= quantile(price_ratio, 0.05, na.rm = TRUE),
         price_ratio <= quantile(price_ratio, 0.95, na.rm = TRUE))
```

**Explanation of non-market rate sales identification methodology:** 
- We compared sale_price to market_value and found a strong overall alignment between the two.
- We examined the price_ratio (sale_price divided by market_value) and identified extreme values, including a maximum around 76, which likely reflects non market transactions or data issues.
- Because these extremes can distort the distribution, we used quantiles to keep only the statistically dominant range of the data.
- Based on this inspection, the 0.01 and 0.99 quantiles fall within an acceptable error range, so we filtered observations to this interval.

#### Adjust variables

```{r}
# Add basic variables
opa_filtered <- opa_filtered %>%
  mutate(age = year(as.Date(sale_date)) - as.numeric(year_built)) %>%
  mutate(bb_ratio = number_of_bathrooms / number_of_bedrooms)

# Pick useful variables
opa_var <- opa_filtered %>%
  dplyr::select(
    objectid, geometry, 
    sale_date, sale_price, market_value, building_code_description_new,
    total_livable_area, number_of_bedrooms, number_of_bathrooms, bb_ratio,
    number_stories, garage_spaces, 
    central_air, quality_grade, view_type,
    interior_condition, exterior_condition, age, 
    off_street_open, zip_code, census_tract, zoning, owner_1,
    category_code_description, fireplaces
  )

```

### Step 2: Load and Clean Secondary Data

#### ACS data

```{r}
acs_vars <- c(
  # Median HH Income
  med_hh_income       = "B19013_001",
  # Housing Cost Burden (Renter)
  burden_renter_total = "B25070_001",
  burden_renter_30    = "B25070_007",  # 30-34.9%
  burden_renter_35    = "B25070_008",  # 35-39.9%
  burden_renter_40    = "B25070_009",  # 40-49.9%
  burden_renter_50    = "B25070_010",  # 50%+
  # Housing Cost Burden (Owner)
  burden_owner_total  = "B25091_001",
  burden_owner_30     = "B25091_008",  # 30-34.9%
  burden_owner_35     = "B25091_009",  # 35-39.9%
  burden_owner_40     = "B25091_010",  # 40-49.9%
  burden_owner_50     = "B25091_011",  # 50%+
  # Poverty
  poverty_total       = "B17001_001",
  poverty_below       = "B17001_002",
  # Housing Units by Type
  total_households    = "B11001_001",
  single_family       = "B25024_002",  # 1-unit detached
  single_family_att   = "B25024_003",  # 1-unit attached
  multi_family        = "B25024_004",  # 2 units
  multi_family_3_4    = "B25024_005",  # 3-4 units
  multi_family_5_9    = "B25024_006",  # 5-9 units
  multi_family_10_19  = "B25024_007",  # 10-19 units
  multi_family_20_49  = "B25024_008",  # 20-49 units
  multi_family_50plus = "B25024_009",  # 50+ units
  # Education (25+)
  edu_total           = "B15003_001",
  edu_bachelor        = "B15003_022",  # Bachelor's degree
  edu_master          = "B15003_023",  # Master's degree
  edu_professional    = "B15003_024",  # Professional degree
  edu_phd             = "B15003_025",  # Doctorate degree
  # Commute Mode
  commute_total       = "B08301_001",
  commute_car         = "B08301_002",  # Car, truck, van
  commute_transit     = "B08301_010"   # Public transportation
)

# Get geometry with just one variable
acs_geom <- get_acs(
  geography = "tract",
  variables = "B19013_001",
  state     = "PA",
  county    = "Philadelphia",
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
) %>%
  dplyr::select(GEOID, geometry)

acs_wide <- get_acs(
  geography = "tract",
  variables = acs_vars,
  state     = "PA",
  county    = "Philadelphia",
  year      = 2023,
  survey    = "acs5",
  geometry  = FALSE
) %>%
  dplyr::select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    burden_renter_rate      = (burden_renter_30 + burden_renter_35 + 
                               burden_renter_40 + burden_renter_50) / burden_renter_total,
    burden_owner_rate       = (burden_owner_30 + burden_owner_35 + 
                               burden_owner_40 + burden_owner_50) / burden_owner_total,
    poverty_rate            = poverty_below / poverty_total,
    single_family_total     = single_family + single_family_att,
    multi_family_total      = multi_family + multi_family_3_4 + multi_family_5_9 +
                              multi_family_10_19 + multi_family_20_49 + multi_family_50plus,
    edu_bachelors_plus_rate = (edu_bachelor + edu_master + 
                               edu_professional + edu_phd) / edu_total,
    commute_car_share       = commute_car / commute_total,
    commute_transit_share   = commute_transit / commute_total
  )

# Join together
acs <- acs_geom %>%
  left_join(acs_wide, by = "GEOID") %>%
  st_as_sf()

# Pick those useful
acs <- acs %>%
  select(GEOID, NAME, geometry,
         med_hh_income,
         burden_renter_rate,
         burden_owner_rate,
         poverty_rate,
         single_family_total,
         multi_family_total,
         total_households,
         edu_bachelors_plus_rate,
         commute_car_share,
         commute_transit_share)

```

#### Spatial amentities

```{r}


```

#### Check the CRS and join to opa

```{r}
```


```{r}
```
## Phase2: Feature Engineering

### Step 1: what

```{r}
opa_path <- "C:/Users/10133/Desktop/UPENN/MUSA5080_PPA/Portfolio/lab_3/opa_properties_public.geojson"
opa <- st_read(opa_path, quiet = FALSE)
```

------------------------------------------------------------------------
